---
title: "p8105_hw3_lah2231"
output: github_document
date: "2024-10-13"
---
```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(lubridate)
```

# Problem 1

```{r}
devtools::install_github("P8105/p8105.datasets")
library(p8105.datasets)
```

```{r}
ny_weather <- ny_noaa
head(ny_weather)
#View(ny_weather)
```
```{r}
str(ny_weather)
colnames(ny_weather)
nrow(ny_weather)
ncol(ny_weather)
```
```{r}
colSums(is.na(ny_weather))
```

The NOAA dataset contains information about weather data such as temperature and snowfall. The columns are called `r colnames(ny_weather)`. The dataset contains `r nrow(ny_weather)` rows and `r ncol(ny_weather)` columns. The variables recorded are for example precipitation, snowfall, snow depth, maximum temperature and minimum temperature. In addition to that there are station identification numbers and dates of observations.

There are `r sum(is.na(ny_weather))` missing values in the dataset overall. Precipitation has `r sum(is.na(ny_weather[,"prcp"]))` missing values, snowfall has `r sum(is.na(ny_weather[,"snow"]))` missing, snow depth has `r sum(is.na(ny_weather[,"snwd"]))` missing, maximum temperature has `r sum(is.na(ny_weather[,"tmax"]))` missing, and minimum temperature has `r sum(is.na(ny_weather[,"tmin"]))` missing.

## Data cleaning
```{r}
ny_weather_clean <- ny_noaa %>%
  mutate(
    year = year(date),
    month = month(date, label = TRUE),
    day = day(date),
    tmax = as.numeric(tmax) / 10,
    tmin = as.numeric(tmin) / 10,
    prcp = prcp / 10
  )
head(ny_weather_clean)
```
I cleaned the data by turning the date into separate variables for year, month and day and I changed the temperature from tenths of degrees celsius to degrees celsius and precipitation from teths of milimeters to mm.

## Common snowfall values
```{r}
snowfall_common <- ny_weather_clean %>%
  filter(!is.na(snow)) %>% 
  group_by(snow) %>%
  summarize(count = n()) %>%
  arrange(desc(count))

head(snowfall_common)

```
The most common snowfall value is zero, which is probably the case because it does not snow in New York on most days.

## Plotting 
```{r}
avg_tmax_simple <- ny_weather_clean %>%
  filter((month == "Jan") | (month == "Jul")) %>%
  filter(!is.na(tmax)) %>%
  group_by(year, month) %>%
  summarize(avg_tmax = mean(tmax)) %>%
  ungroup()

ggplot(avg_tmax_simple, aes(x = year, y = avg_tmax)) +
  geom_line() +
  facet_wrap(~month, ncol = 1) +
  labs(
    title = "Aerage max temperature",
    x = "year",
    y = "average max temperature"
  )
```
I have made a two panel plot by taking the average mx temp in january and july after removing missing tmax values. One can see that it correlates that in some years if january was rather warm july was also rather warm.

```{r}
ggplot(ny_weather_clean, aes(x = tmin, y = tmax)) +
  geom_bin2d() +
  labs(
    title = "tmax and tmin",
    x = "min temp",
    y = "max temp"
  )
```

Panel 1 shows the relationship between tmax and tmin using a 2D bin plot using geom_bin2d because of the large amount of data.

```{r}

# Panel 2: Distribution of snowfall > 0 and < 100 by year using boxplot
ny_weather_clean %>%
  filter(snow > 0 & snow < 100) %>%
  ggplot(aes(x = factor(year), y = snow)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Snowfall (0 < Snowfall < 100 mm) by Year",
    x = "Year",
    y = "Snowfall (mm)"
  ) +
  theme_minimal()
```





































































