---
title: "p8105_hw3_lah2231"
output: github_document
date: "2024-10-13"
---

# Problem 1

```{r}
devtools::install_github("P8105/p8105.datasets")
library(p8105.datasets)
```

```{r}
ny_weather <- ny_noaa
head(ny_weather)
```
```{r}
str(ny_weather)
colnames(ny_weather)
nrow(ny_weather)
ncol(ny_weather)
```
```{r}
colSums(is.na(ny_weather))
```

The NOAA dataset contains information about weather data such as temperature and snowfall. The columns are called `r colnames(ny_weather)`. The dataset contains `r nrow(ny_weather)` rows and `r ncol(ny_weather)` columns. The variables recorded are for example precipitation, snowfall, snow depth, maximum temperature and minimum temperature. In addition to that there are station identification numbers and dates of observations.

There are `r sum(is.na(ny_weather))` missing values in the dataset overall. Precipitation has `r sum(is.na(ny_weather[,"prcp"]))` missing values, snowfall has `r sum(is.na(ny_weather[,"snow"]))` missing, snow depth has `r sum(is.na(ny_weather[,"snwd"]))` missing, maximum temperature has `r sum(is.na(ny_weather[,"tmax"]))` missing, and minimum temperature has `r sum(is.na(ny_weather[,"tmin"]))` missing.










































































